Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-01 paper/paper.qmd
add discussion

---
title: "Understanding the Impact of Mortgage State and Income on Household Poverty Status"
subtitle: "An Analysis of Household Poverty Status in the United States in 2019"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households' total annual earning less than $50000 and for renters compared to property owners. All households with incomes exceeding $100000 are not in poverty. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Over the last few decades, addressing poverty has remained a persistent issue in the United States, leading to the implementation of multiple policies in an effort to mitigate this problem [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession, marking the last full year before the onset of COVID-19 pandemic. The robustness of the U.S. economy during this period was evidenced by low poverty rates, indicating a notable decrease in economic hardship characterized by low income. The poverty rate in the United States was estimated to be 10.5% in 2019, down from 11.8% in 2018, marking the fifth consecutive year of decline [@Semega_2022]. Despite various interventions and economic policies, understanding the intricate dynamics influencing household poverty status is still difficult. Therefore, it is important for us to analyze why some US household are in poverty in 2019 and how various demographic and financial factors may has an influence on it. Perhaps then we could understand how to reduce the number of households in poverty. We analyze the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. Supplemental Poverty Measure was developed due to the continued criticism towards the official poverty measure in the US. SPM addresses numerous concerns of official measure critics, and its intent is to provide an improved statistical picture of poverty [@Gesumaria_2015]. This allows us to discover trends and address potential economic and societal situations that may cause the poverty status for some US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various demographic and financial factors captured in the SPM dataset. We selected data features: mortgage state and household annual total income. The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is trained using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCB.

@@

## Estimand
The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is trained using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCS.

@@

## Data Source
This report uses the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset provided by the United States Census Bureau as our main source of data. The Census Bureau provides information about the people and economy of the United States, with a goal to support economic growth, enhance scientific knowledge, and assist in making informed decisions. The official poverty measure has been criticized due to its multiple limitations. For example, it does not consider benefits from most of the largest programs that aid the low-income population. For instance, it uses money income before taxes, so it does not measure the income available for individuals to spend, which for most people is after-tax income. Therefore, any effects of tax credits designed to assist persons with low income are not considered by the official measure. The Supplemental Poverty Measure (SPM), similar to the official poverty measure, tackles the shortcomings of poverty assessment. It determines poverty status by gauging resources against a defined standard of living. This standard considers expenditures on essentials like food, clothing, shelter, and utilities, plus a bit more for additional expenses. The resources assessed include disposable income, accounting for taxes and some non-cash benefits, available to cover these needs [@Gesumaria_2015]. In 2019, the dataset contained a total of 157,959 entries.

## Features
The original SPM 2019 dataset, which shows in @tbl-raw, contains 157959 data entries and many variables. Since it is difficult to observe such a large dataset, this report will only explore and analyze through several data features. We first chose these 17 variables: h_seq, spm_poor, spm_tenmortstatus, spm_totval, spm_snapsub, spm_caphousesub, spm_schlunch, spm_engval, spm_wicval, spm_fedtax, spm_eitc, spm_actc, spm_fica, spm_sttax, spm_childsuppd, spm_capwkccxpns, and spm_medxpns. This is because we want to calculate the income after tax, plus all the subsidies and benefits, minus all the expenses, just like how the SPM does it. If we only choose income before tax to analyze, then our analysis will be like the official poverty measure, which requires improvement.

@@

@tbl-summary is a summary of the cleaned data, showing detailed statistics about the dataset. As we can see from the table, there are more households which are not in poverty. Respondents cover a wide range of income levels, with "50k-100k" and "100k-150k" being more heavily represented. Also, the mortgage states of the households vary, each of the three mortgage states being fairly evenly represented.

@@

@fig-income shows the relationship between household annual total income and their poverty status. We can see that among households earning more than 100k annually, no household is in poverty. For household with income between 50k to 100k, almost all of them are not in poverty with only a very small number of households in poverty. Within the income range of 5k-50k, the majority of households are not in poverty, but there are some families in poverty. However, for households earning below 5k annually, almost all of them are in poverty. This indicates a strong positive correlation between higher income levels and financial stability.

```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = income, fill = poverty_status)) +
  geom_bar(position = "dodge", width = 0.5) +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Income level",
    y = "Number of households",
    fill = "Poverty status"
  )
```

@@

modelsummary(
  list(
    "In poverty" = poverty_prediction_model
  ),
  statistic = "mad"
  )
```

@@

## Relationship betwen Mortgage State and Poverty Status

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Relationship betwen Income and Poverty Status

Average annual expenditure $63,036

>>>>

---
title: "Understanding the Impact of Mortgage State and Income on Household Poverty Status"
subtitle: "An Analysis of Household Poverty Status in the United States in 2019"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level after tax and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households' total annual earning less than $50000 and for renters compared to property owners. All households with incomes exceeding $100000 are not in poverty. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Over the last few decades, addressing poverty has remained a persistent issue in the United States, leading to the implementation of many policies in an effort to mitigate this problem [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession. The robustness of the U.S. economy during this period was evidenced by low poverty rates, indicating a notable decrease in economic hardship. The poverty rate in the United States was estimated to be 10.5% in 2019, down from 11.8% in 2018, marking the fifth consecutive year of decline [@Semega_2022]. In order to reduce poverty further, it is important for us to analyze why some US household are in poverty in 2019 and how various demographic and financial factors may has an influence on it. We analyze the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of official poverty measure critics, and its intent is to provide an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. This allows us to discover trends and address potential financial and societal situations that may cause the poverty status for some US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various demographic and financial factors captured in the SPM dataset. We select 17 data features for analysis. After some data cleaning, we summarize 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus benefits and minus expenses. The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is trained using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCS.

@@

@@

## Data Source
This report uses the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset provided by the United States Census Bureau as our main source of data. The Census Bureau provides information about the people and economy of the United States, with a goal to support economic growth, enhance scientific knowledge, and assist in making informed decisions. The official poverty measure has been criticized due to its multiple limitations [@sgp2019]. For example, it does not consider benefits from most of the largest programs that aid the low-income population. For instance, it uses money income before taxes, so it does not measure the income available for individuals to spend, which for most people is after-tax income. Therefore, any effects of tax credits designed to assist persons with low income are not considered by the official measure. The Supplemental Poverty Measure (SPM), similar to the official poverty measure, tackles the shortcomings of poverty assessment. It determines poverty status by gauging resources against a defined standard of living. This standard considers expenditures on essentials like food, clothing, shelter, and utilities, plus a bit more for additional expenses. The resources assessed include disposable income, accounting for taxes and some non-cash benefits, available to cover these needs [@Gesumaria_2015]. In 2019, the dataset contained a total of 157,959 entries.

## Features
The original SPM 2019 dataset, which shows in @tbl-raw, contains 157959 data entries and many variables. Since it is difficult to observe such a large dataset, this report will only explore and analyze through several data features. We first chose these 17 variables: h_seq, spm_poor, spm_tenmortstatus, spm_totval, spm_snapsub, spm_caphousesub, spm_schlunch, spm_engval, spm_wicval, spm_fedtax, spm_eitc, spm_actc, spm_fica, spm_sttax, spm_childsuppd, spm_capwkccxpns, and spm_medxpns. This is because we want to calculate the income after tax, plus all the subsidies and benefits, minus all the expenses, just like how the SPM does it [@sgp2019]. If we only choose income before tax to analyze, then our analysis will be like the official poverty measure, which requires improvement.

@@

@tbl-summary is a summary of the cleaned data, showing detailed statistics about the dataset. As we can see from the table, there are more households which are not in poverty. Respondents cover a wide range of income levels, with "5k-50k" and "50k-100k" being more heavily represented. Also, the mortgage states of the households vary, each of the three mortgage states being fairly evenly represented.

@@

@fig-income shows the relationship between household annual total income and their poverty status. We can see that among households earning more than 100k annually, no household is in poverty. For household with income between 50k to 100k, almost all of them are not in poverty with only a very small number of households in poverty. Within the income range of 5k-50k, the majority of households are not in poverty, but there are some families in poverty. However, for households earning below 5k annually, all of them are in poverty. This indicates a strong positive correlation between higher income levels and financial stability.

```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false

poverty2019 |>
  ggplot(aes(x = income, fill = poverty_status)) +
  geom_bar(position = position_dodge2(width = 0.5, preserve = "single")) +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Income level",
    y = "Number of households",
    fill = "Poverty status"
  )
```

@@

modelsummary(
  list(
    "In poverty" = poverty_prediction_model
  ),
  statistic = "mad"
  )
```
\newpage

@@

## Relationship betwen Mortgage State and Poverty Status
In the USA in 2019, housing costs accounted for more than 30% of household expenditure, making mortgage status a critical factor in determining a family's risk of falling into poverty [@Statistics_2020]. With 11.8% of the total population (38.3 million people) living below the poverty line, it's evident that housing affordability significantly impacts economic stability. The percentage of individuals below the poverty threshold, adjusted for housing costs are: $29,080 for a family of four with a mortgage, $24,413 for a family of four without a mortgage, and $29,194 for a family of four [@Khattar_Pathak_Schweitzer_Khan_McManus_Jones_Benson_Kelly_2023]. From this, we can assumed that families without mortgages experienced the lowest housing costs, leaving more funds available for other essentials and eliminating the need for alternative housing options. In contrast, families with mortgages may face burdensome monthly payments, especially if they have chosen a floating interest rate, making them vulnerable to fluctuations in interest rates. Renters also encounter challenges, as rent increases can strain finances and necessitate either accepting higher housing costs or undertaking the time-consuming and costly process of finding new accommodations. These housing-related challenges exacerbate the risk of poverty for affected families, underscoring the importance of addressing housing affordability as a key component of poverty alleviation efforts.

## Relationship betwen Income and Poverty Status
The relationship between a household's income after tax and its poverty status is characterized by a negative correlation, wherein higher income corresponds to a reduced probability of experiencing poverty. This correlation shows the role of financial resources in mitigating economic hardship and facilitating socio-economic well-being. However, in the United States, the cost of living is notably high, low-income households often struggle to meet basic needs, resulting in an unending cycle of poverty. In 2019, the poverty guideline for a one-person household was set at $12,490 annually [@ASPE]. So, it is evident that families earning less than $5,000 are definitely living below the poverty line. Conversely, individuals with high incomes are less likely to experience poverty, as their financial resources enable them to comfortably cover living expenses. They will be able to have savings accumulation, further consolidating their financial security and contributing to wealth accumulation over time.

For household in poverty, escaping poverty is tough, but the US government and communities are stepping up to help. For example, single parents often struggle with low incomes. But they can get help from government programs such as the Supplemental Nutrition Assistance Program (SNAP) [@snap], commonly known as food stamps, which provides eligible low-income individuals and families with funds to purchase nutritious food. They can also receive child support payments to help cover expenses for their kids. Families with low incomes and lots of kids can also get help. Programs like the Child Tax Credit (CTC) [@ctc], which provides tax credits to eligible families with qualifying children. Other programs, like housing assistance and Medicaid, provide support with things like affordable housing and healthcare. These programs are all aimed at helping families in need to get back on their feet and improve their poverty situation.
